Structuring College Access: The Market Segment Model and College Board Geomarkets

### 2025-04-25 text/front_end_v4.qmd
created front end v4

On January 9, 2019 a public research university purchased College Board *Student Search Service* order no. $448922$, thereby obtaining the contact information of out-of-state prospective students, who would subsequently receive brochures, postcards, email, and targeted social media. This was a large student list purchase, yielding $122,426$ "names." At $\$0.45$ per name, the order cost $\$55,091.70$. The order specified three "search filters" to control which names were purchased: first, the 2020 high school graduating class; second, under the heading "College Board Exams," PSAT score range of 1070 - 1180; and third, under the heading "Geography", the order filtered on 67 "Geomarkets," identified by titles like "IL08 - Northwest Suburbs" and "IL09 - North Shore."^[On the same day, the university purchased two additional orders that filtered on PSAT score ranges of $1190 - 1260$ and $1270 - 1520$, respectively, but selected the same HS graduating class and Geomarkets.] The Geomarket filter carves states and metroplitan areas into smaller geographic units. Upon closer inspection, this student list purchase selected all Geomarkets in the Chicago-land area except for "IL10 - City of Chicago," which is significantly poorer and less White.

@@

Geomarkets carve states and metroplitan areas into smaller geographic units. Upon closer inspection, the three orders selected on all Geomarkets in the Chicago-land area except for "IL10 - City of Chicago", which is significantly poorer and less White. In the DC/Maryland/Virginia metro area, the orders selected the affluent "MD02 - Montgomery Metropolitan," "MD03 - Central Maryland (w/o Baltimore)," "VA01 - Arlington & Alexandria," and "VA02 - Fairfax County" Geomarkets, but excluded the "MD05 - Prince Georges Metropolitan," "MD07 - Baltimore (Urban)," and "DC01 - District of Columbia" Geomarkets, which are poorer and have a higher share of Black residents. An analysis of 500 \[check\] student lists purchased by 9 public research universities found that 11% of purchases filtered on Geomarket [@RN5012]. Recent research suggests that student lists are surprisingly important for college access outcomes and college graduation outcomes, particularly for first-generation and underrepresented minority students [@RN4739; @RN4894]. To the extent that Geomarket borders are correlated with race and class, the Geomarket selections observed here raise concern that Geomarkets -- and other geographic search filters -- may systematically filter working class and non-white students from educational opportunities.
-->
Sociology offers several perspectives to describe the process of sorting students into colleges. The status attainment model argues that student postsecondary destination is a function of family socioeconomic background, particularly parental education [e.g., @RN915; @RN1273; @RN1268; @RN5033; @RN1300]. The cultural capital model argues that bourgeois households bestow the pedigree, social networks, and information necessary for their progeny to claim spots at selective colleges [@RN1228; @RN1241; @RN4991] [ADD CITES]. @RN4873 describes schools as "sorting machines" that create, incorporate, and assign students to categories that determine educational opportunities, often reinforcing reinforcing racial and class inequality. Finally, scholarship on enrollment management offers an agentic depiction of organizational behavior, showing that colleges expend considerable resources recruiting desirable prospects [@RN3519; @RN4758; @cottom2017lower; @RN4324]. All of these perspectives share the view that transitions from school to college tend towards social reproduction. We argue that these perspectives have ignored an important mechanism of social reproduction: third-party vendors create products that sort students on behalf of college, often utilizing historical snapshots of student demand to make recommendations about where colleges should prioritize recruiting efforts. While third-party recruiting products have become increasingly powerful, granular, and ubiquitous in the era of platform capitalism, we argue that a watershed moment in their develeopment was the creation of Geomarkets by the College Board and UPenn professor, Robert Zemsky.

In 1983, Robert Zemsky and Penny Odell authored *The Structure of College Choice*. College Board published the book, provided data, and financed the project, as part of their "efforts to help colleges estimate their enrollment potential, efforts which had faltered largely because the smallest geographic unit used in these analyses was the state" [@RN4982, p. x]. @RN4982 developed Geomarkets and the Market Segment Model as an effort to "capture and quantify" (p. 11) the knowledge of admissions officers, based on the idea that "a good recruiter knows where to look for prospective applicants, as seen in the students' willingness or reluctance to travel" (p. 11).  Created by analyzing the SAT score-sending behavior of 1980 high school seniors, the thesis of the Market Segment Model is that student demand for your institution is a function of social class and geography. Therefore, colleges should recruit from territories that contain large populations of your target social class. Geomarkets are geographic borders meant to define local recruiting markets, a territory of an admissions recruiter. Figure X, panel A shows Geomarkets in the Chicago-land area. The Market Segment Model predicts how demand for a particular college varies by Geomarket, based on characteristics of households in the Geomarket.

Geomarkets have been incorporated into the supply-side structure of college access in several ways. Albeit anecdotally, Geomarkets became an organizing principle for how college admissions offices allocate admissions recruiters to territories. Figure X panel B shows that Chicago-land Geomarkets define recruiting territories for University of Chicago admissions recruiters (Appendix Figures BLAH BLAH). Second, Geomarkets are the basis for the College Board Enrollment Planning Service (EPS). Founded in 1984 and still active today, EPS software recommends which Geomarkets a college should recruit from and which schools/communities they should prioritize within targeted Geomarkets. According to @RN5021, in 1995, 37% of 4-year public institutions and 49% of 4-year private institutions used EPS, while 41% of 4-year publics and 16% of 4-year privates used ACT's market analysis product, which was based on EPS.

@@

Scholarship at the nexus of college access and recruiting has focused on the behavior of colleges [e.g., @cottom2017lower; @RN4758; @RN4759; @RN3519], contributing to the belief that recruiting is a function of individual colleges. Scholarship has failed to investigate the broader enrollment management industry -- consisting of third-party vendors, their products and consulting solutions -- as a set of mechanisms that structure college access. These mechanisms, which lie upstream of the behavior of individual colleges, are incorporated by individual colleges in ways that funnel certain kinds of students to certain kinds of institutions. This omission is surprising because sociology is concerned with sorting [@RN4873], particularly sorting into college.

@@

Our analyses address the two research questions, which speak to how Geomarkets are utilized within EPS and within the Student Search Service student list product. First, what is the socioeconomic and racial variation between geomarkets in metropolitan areas and how does this variation change over time? We address this question by spatially joining the geomarket shapefile to Census data about socioeconomic and racial characteristics. Second, how does the socioeconomic and racial composition of included versus included prospects vary when student list purchases filter on particular geomarkets? We address this question by analyzing actual student list purchases that utilize commonly used search filters (e.g., PSAT score, GPA) but do not filter on Geomarkets. We simulate which prospects in particular metropolitan areas would have been included/excluded had the student list purchase filtered on particular Geomarkets. We obtained these data by issuing public records requests to public universities.

>>>>

On January 9, 2019 a public research university purchased College Board *Student Search Service* order no. $448922$, thereby obtaining the contact information of out-of-state prospective students, who would subsequently receive brochures, postcards, email, and targeted social media. This was a large student list purchase, yielding $122,426$ "names." At $\$0.45$ per name, the order cost $\$55,091.70$. The order specified three "search filters" to control which names were purchased: first, the 2020 high school graduating class; second, under the heading "College Board Exams," PSAT score range of 1070 - 1180; and third, under the heading "Geography", the order filtered on 67 "Geomarkets," identified by titles like "IL08 - Northwest Suburbs" and "IL09 - North Shore."^[On the same day, the university purchased two additional orders that filtered on PSAT score ranges of $1190 - 1260$ and $1270 - 1520$, respectively, but selected the same HS graduating class and Geomarkets.] The Geomarket filter carves states and metroplitan areas into smaller geographic units. Upon closer inspection, this student list purchase selected all Geomarkets in the Chicago-land area except for "IL10 - City of Chicago," which is significantly poorer and has a higher share of Black residents. Similarly, in the DC/Maryland/Virginia metro area, the order selected the affluent "MD02 - Montgomery Metropolitan," "MD03 - Central Maryland (w/o Baltimore)," "VA01 - Arlington & Alexandria," and "VA02 - Fairfax County" Geomarkets, but excluded the "MD05 - Prince Georges Metropolitan," "MD07 - Baltimore (Urban)," and "DC01 - District of Columbia" Geomarkets.

@@

Geomarkets carve states and metroplitan areas into smaller geographic units. Upon closer inspection, the three orders selected on all Geomarkets in the Chicago-land area except for "IL10 - City of Chicago", which is significantly poorer and less White. In the DC/Maryland/Virginia metro area, the orders selected the affluent "MD02 - Montgomery Metropolitan," "MD03 - Central Maryland (w/o Baltimore)," "VA01 - Arlington & Alexandria," and "VA02 - Fairfax County" Geomarkets, but excluded the "MD05 - Prince Georges Metropolitan," "MD07 - Baltimore (Urban)," and "DC01 - District of Columbia" Geomarkets, which are poorer and have a higher share of Black residents. An analysis of 500 \[check\] student lists purchased by 9 public research universities found that 11% of purchases filtered on Geomarket [@RN5012]. Recent research suggests that student lists are surprisingly important for college access outcomes and college graduation outcomes, particularly for first-generation and underrepresented minority students [@RN4739; @RN4894]. To the extent that Geomarket borders are correlated with race and class, the Geomarket selections observed here raise concern that Geomarkets -- and other geographic search filters -- may systematically filter working class and non-white students from educational opportunities.
-->
Sociology offers several perspectives to describe the process of sorting students into colleges as a process that tends toward social reproduction [@RN4873; @RN2377]. On the student demand side, the status attainment model argues that student postsecondary destination is a function of family socioeconomic background, particularly parental education [e.g., @RN915; @RN1273; @RN1268; @RN5033; @RN1300]. The cultural capital model argues that bourgeois households bestow the pedigree, social networks, and information necessary for their progeny to claim spots at selective colleges [@RN1228; @RN1241; @RN4991] [ADD CITES]. @RN4873 describes schools as "sorting machines" that create, incorporate, and assign students to categories that determine educational opportunities, often reinforcing reinforcing racial and class inequality. On the supply-side, the credentialism literature recognizes that colleges have a financial incentive for educational credentials to be required to claim advantageous social and economic positions [@RN1297], and this race for credentials benefits affluent households [@RN861] [cite others]. Scholarship on enrollment management offers an agentic depiction of organizational behavior, showing that colleges expend considerable resources recruiting desirable prospects [@RN3519; @RN4758; @cottom2017lower; @RN4324].

We argue that these perspectives ignore an important mechansim of social reproduction: third-party vendors create products that sort students on behalf of college. These products often amplify historic inequality in educational opportunity by using past snapshots of student demand to make recommendations about where colleges should prioritize recruiting efforts. In the contemporary era of private equity funded platform capitalism, third-party recruiting products have become increasingly powerful, granular, and ubiquitous. However, we argue that a watershed moment in their develeopment was the creation of Geomarkets by the College Board and UPenn professor, Robert Zemsky.

In 1983, Robert Zemsky and Penny Odell authored *The Structure of College Choice*. College Board underwrote the project -- providing data and funding -- and published the book as part of their nascent efforts "to help colleges estimate their enrollment potential, efforts which had faltered largely because the smallest geographic unit used in these analyses was the state" [@RN4982, p. x]. The broader project was conceptualized as an effort to to "capture and quantify" (p. 11) the knowledge of admissions officers, based on the idea that "a good recruiter knows where to look for prospective applicants, as seen in the students' willingness or reluctance to travel" (p. 11). @RN4982 developed the Market Segment Model by analyzing the SAT score-sending behavior of 1980 high school seniors. The thesis of the Market Segment Model is that student demand for your institution is a function of social class and geography. Therefore, colleges should recruit from territories that contain large populations of your target social class. Geomarkets are geographic borders meant to define local recruiting markets, a territory of an admissions recruiter. Figure X, panel A shows Geomarkets in the Chicago-land area. As a planning tool for colleges, the Market Segment Model predicts how demand for a particular college varies by Geomarket, based on characteristics of households in the Geomarket.

Geomarkets have been incorporated into the supply-side structure of college access in three ways. Anecdotally, Geomarkets became an organizing principle for how college admissions offices allocate admissions recruiters to territories. Figure X panel B shows that Chicago-land Geomarkets define recruiting territories for University of Chicago admissions recruiters (Appendix Figures BLAH BLAH). Second, Geomarkets are the basis for the College Board Enrollment Planning Service (EPS), which was in 1984 and remains active today. EPS software recommends which Geomarkets a college should recruit from and which schools/communities they should prioritize within targeted Geomarkets. According to @RN5021, in 1995, 37% of 4-year public institutions and 49% of 4-year private institutions used EPS, while 41% of 4-year publics and 16% of 4-year privates used ACT's market analysis product, which was based on EPS.

@@

Scholarship on enrollment management has focused on college behavior [e.g., @cottom2017lower; @RN4758; @RN4759; @RN3519], contributing to the belief that recruiting is a function of individual colleges. Scholarship has failed to investigate the broader enrollment management industry -- consisting of third-party vendors, their products and consulting solutions -- as a set of mechanisms that structure college access. These mechanisms, which lie upstream of the behavior of individual colleges, are incorporated by individual colleges in ways that funnel certain kinds of students to certain kinds of institutions. This omission is surprising because sociology is concerned with sorting [@RN4873], particularly sorting into college.

@@

Our analyses address the two research questions, which speak to how Geomarkets are utilized within EPS and within the Student Search Service student list product. First, what is the socioeconomic and racial variation between geomarkets in metropolitan areas and how does this variation change over time? We address this question by spatially joining the geomarket shapefile to Census data about socioeconomic and racial characteristics. Second, how does the socioeconomic and racial composition of included versus included prospects vary when student list purchases filter on particular geomarkets? This research question is motivated by @RN4982, which recommends that selective colleges target affluent Geomarkets. Additionally, prior research on enrollment management behavior finds that selective colleges focus recruiting visits on high schools in affluent, predominantly white communities [@RN3519; @RN4758]. We analyze this research question using data from actual student list purchases, which were collected using public records requests. We analyze student list purchases that include all Geomarkets in a metropolitan area in only to assess which prospects would be included/excluded by had the student list purchase filtered on particular Geomarkets.
